### 2024-08-11 doc/Thesis/chapter/021-theory-calculations.tex
nearly finish with todos from todo.md

But this does not mean that computers are more intelligent than humans or \linebreak[4] sentient\cite{searle1980,searle1999married} even though some claim\cite{tiku2022google} and it certainly does not mean that research is over. 
AI is still a young field, which is strongly growing and is gaining ubiquitous status. 
It is slowly creeping into every aspect of modern human life just like electricity around one hundred years ago. 
Realms in which AI is gaining traction are: 
%
playing board games (and beating humans)\cite{Silver2017,Feng1999,Campbell2002}, 
image recognition (very popular for medical diagnosis)\cite{Li2020,Deo2015,Topol2019,Fujiyoshi2019}, 
chemistry\cite{Westermayr2019,goh2017chemception,jha2018elemnet}, 
cyber security\cite{Sarker2021},
facial recognition (to prevent theft of lavatory paper at public toilet in China\cite{Andrews2017}),
financial sector (as robo-advisors)\cite{Littman2021},
natural language processing (NLP)\cite{Koroteev2021,Liu2021gpt,Parviainen2021} 
(which can also create code and pictures through scalable vector graphics (SVG))
and even creative tasks like 
creating non existing faces\cite{Mansourifar2020}, 
create graphic artwork (DALL-E 2)\cite{Marcus2022} or 
making video games\cite{Guzdial2016}.
%
It is nearly hard to find a field where \gls{ai} is not used in some way. 
This steady incorporation of \gls{ai} leads to the so called \gls{ai} effect\cite{McCorduck1982,ai100}: 
certain fields get incorporated into \gls{ai} research and practice,
such that after some time of general use it is no more considered \gls{ai} (e.g.\ spam filter or web searches).
Google CEO Sundar Pichai even goes as far and said: 
"AI is one of the most important things humanity is working on. It is more profound than [...] electricity or fire"\cite{Hassan2020}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{EMMA}
\subsection{Evolutionary Model-based Multiresponse Approach}
\Gls{emma} is an implementation of \gls{pso} in the \texttt{R} programming language. 
Each time step the dependent variables for all possible input variable combinations %$\mathbf{x} = \{x_1,\cdots ,x_N\}$ 
are predicted with the help of \gls{mars}. 
The \gls{mars} \gls{rf} is then used to chose the next position for each particle.
\Gls{mars} is a regression method introduced by Friedman's 1991 paper 
"Multivariate Adaptive Regression Splines"\cite{friedman1991multivariate}. 
The \texttt{CRAN} (Comprehensive R Archive Network) package which implements \gls{mars} is called \texttt{earth} due to \gls{mars} being trade marked\cite{mars}.
%%% MARS 
\gls{mars} is an extension of multivariate linear regression.
Friedman presented \gls{mars} as an alternative to piecewise polynomials (splines) and local averaging methods (e.g.\ kernel functions). %, whose number of parameter quickly \td{exceeds the practical} for moderate dimensional problems ($n>5$). 
Advantages of \Gls{mars} are that it uses less parameters and it produces continuous models with continuous derivatives.
%
%to higher dimensions (n > 2) is straightforward in principle but difficult in practice. 
%"These difficulties are related to the so-called “curse-of-dimensionality”, a phrase coined by Bellman (1961) to express the fact that exponentially increasing numbers of points are needed to densely populate Euclidean spaces of increasing dimension."\cite{friedman1988fitting}
%
Furthermore, \gls{mars} has the ability to fit (possibly complex) interactions and non-linearities without losing interpretability. 
%
%- "The key ingredient that advances this approach to general settings is the ability to fit (possibly complex) interactions among the variables through the product terms that are permitted to enter the approximation (9), if required by the fit."\cite{friedman1988fitting}
%% MARS MODEL SELECTION
%- "The approximation is developed in a forward/backwards stepwise recursive manner in analogy with the recursive partitioning approach. "\cite{friedman1988fitting}
%

@@

%FULL FACTORIAL
A naive approach to an experiment design is the full factorial \gls{doe}.
Each possible combination of discrete values is tested.
The full factorial design will mostly be infeasible due the curse of dimensionality\cite{cherkassky1998learning} despite it being the most informative design.
% 2-LVL FACTORIAL
The 2-level factorial design provides an alternative with $2^d$ experiments (where $d$ is the number of independent variables). 
Drawbacks of 2-level factorial \gls{doe}s include no data about the inside of the search space and infeasibility for high dimensional problems.
In a full factorial or 2-level factorial design most experiments are redundant and most resources will be spent exploring high-order interaction effects\cite{gunst2009fractional}, which are often minimal to non-existent.
% RANDOM 
In order to overcome these obstacles a certain number of experiments can be chosen randomly from the search space. 
When a subset is chosen from the factorial design, it is called a fractional factorial design. 
% PLACKETT_BURMAN
The \gls{pb}\cite{vanaja2007design,miller2001using,wang1995hidden} design is a special case of 2-level fractional factorial design, 
where the number of needed experiments $n$ is $n<d+4$ 
(more precisely $n=(\lfloor d\div4\rfloor+1)\cdot4$, where $\lfloor x\rfloor$ denotes the floor function on $x$).
The \gls{pb} design ensures that each combination of levels for any pair of factors appears the same number of times. 
% HAMMERSLEY 
A drawback of 2-level factorial (incl. \gls{pb}) and random fractional designs is that the sample set is likely not evenly distributed across the search space\cite{viana2016tutorial}. 
The Hammersley design\cite{viana2016tutorial,diwekar1997efficient} is based on the Hammersley sequence and produces space filling data points. 
% LATIN HYPER CUBE
The Latin hypercube \gls{doe}\cite{viana2016tutorial,diwekar1997efficient} is a type of orthogonal \gls{doe}, 
which has the feature that each level for each variable will be tested only once. 
A Latin hypercube \gls{doe} can be also created such that data points distribute more uniformly over the search space. 
Latin hypercube \gls{doe}s are mainly used in computer simulations which are purely deterministic and therefore are very precise.
%
%Continuous state spaces must be accommodated by arbitrary discretization.
%\cite{cohn1996neural}
\iffalse
%
\td{For any 2-level factorial DOE applies: "The effect of any factor main effect (A, B, C) or interaction (AB, AC, BC, ABC) is the difference of two averages, the average of the responses correpsonding to +1 levels and the average of the responses corresponding to -1 levels. " - from Gunst2009\cite{gunst2009fractional}}

@@

%%% FEATURE SELECTION 
% https://stats.stackexchange.com/questions/39243/how-does-one-interpret-svm-feature-weights
% gene (feature) selection with SVM https://link.springer.com/content/pdf/10.1023/A:1012487302797.pdf
% Intro to feature selection https://axon.cs.byu.edu/Dan/778/papers/Feature%20Selection/guyon2.pdf
\iffalse
%%% VALIDATION 
\subsection{Validation}
Validation is an important part of data analysis as it allows to estimate model errors. 
Often used methods are analysis of residual error, cross validation and leave-one-out validation\cite{kartam1997artificial}.
\td{to do read kartam}
statistical methods to bewerten a model are: 
cross validation, leave-one-out validation ( special case of cross validation), correlation coefficient for linear fits, mean square error, mean absolute erro
\td{cite from Hawkins 2004\cite{hawkins2004overfitting}}

>>>>

But this does not mean that computers are more intelligent than humans or \linebreak[4] sentient\cite{searle1980,searle1999married} even though some claim\cite{tiku2022google} and it certainly does not mean that research is over. 
\Gls{ai} is still a young field, which is strongly growing and is gaining ubiquitous status. 
It is slowly creeping into every aspect of modern human life just like electricity around one hundred years ago. 
Realms in which AI is gaining traction are: 
%
playing board games (and beating humans)\cite{Silver2017,Feng1999,Campbell2002}, 
image recognition (very popular for medical diagnosis)\cite{Li2020,Deo2015,Topol2019,Fujiyoshi2019}, 
chemistry\cite{Westermayr2019,goh2017chemception,jha2018elemnet}, 
cyber security\cite{Sarker2021},
facial recognition (to prevent theft of lavatory paper at public toilet in China\cite{Andrews2017}),
financial sector (as robo-advisors)\cite{Littman2021},
natural language processing (NLP)\cite{Koroteev2021,Liu2021gpt,Parviainen2021} 
(which can also create code and pictures through scalable vector graphics (SVG))
and even creative tasks like 
creating non existing faces\cite{Mansourifar2020}, 
create graphic artwork (DALL-E 2)\cite{Marcus2022} or 
making video games\cite{Guzdial2016}.
%
It is nearly hard to find a field where \gls{ai} is not used in some way. 
This steady incorporation of \gls{ai} leads to the so called \gls{ai} effect\cite{McCorduck1982,ai100}: 
certain fields get incorporated into \gls{ai} research and practice,
such that after some time of general use it is no more considered \gls{ai} (e.g.\ spam filter or web searches).
Google CEO Sundar Pichai even goes as far and said: 
"AI is one of the most important things humanity is working on. It is more profound than [...] electricity or fire"\cite{Hassan2020}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{EMMA}
\subsection{Evolutionary Model-Based Multiresponse Approach}
\Gls{emma} is an implementation of \gls{pso} in the \texttt{R} programming language. 
Each time step the dependent variables for all possible input variable combinations %$\mathbf{x} = \{x_1,\cdots ,x_N\}$ 
are predicted with the help of \gls{mars}. 
The \gls{mars} \gls{rf} is then used to chose the next position for each particle.
\Gls{mars} is a regression method introduced by Friedman's 1991 paper 
"Multivariate Adaptive Regression Splines"\cite{friedman1991multivariate}. 
The \texttt{CRAN} (Comprehensive R Archive Network) package which implements \gls{mars} is called \texttt{earth} due to \gls{mars} being trade marked\cite{mars}.
%%% MARS 
\gls{mars} is an extension of multivariate linear regression.
Friedman presented \gls{mars} as an alternative to piecewise polynomials (splines) and local averaging methods (e.g.\ kernel functions). %, whose number of parameter quickly \td{exceeds the practical} for moderate dimensional problems ($n>5$). 
Advantages of \Gls{mars} are that it uses less parameters and it produces continuous models with continuous derivatives.
%
%to higher dimensions (n > 2) is straightforward in principle but difficult in practice. 
%"These difficulties are related to the so-called “curse-of-dimensionality”, a phrase coined by Bellman (1961) to express the fact that exponentially increasing numbers of points are needed to densely populate Euclidean spaces of increasing dimension."\cite{friedman1988fitting}
%
Furthermore, \gls{mars} has the ability to fit (possibly complex) interactions and non-linearities without losing interpretability. 
%
%- "The key ingredient that advances this approach to general settings is the ability to fit (possibly complex) interactions among the variables through the product terms that are permitted to enter the approximation (9), if required by the fit."\cite{friedman1988fitting}
%% MARS MODEL SELECTION
%- "The approximation is developed in a forward/backwards stepwise recursive manner in analogy with the recursive partitioning approach. "\cite{friedman1988fitting}
%

@@

%FULL FACTORIAL
A naive approach to an experiment design is the full factorial \gls{doe}.
Each possible combination of discrete values is tested.
The full factorial design will mostly be infeasible due the curse of dimensionality\cite{cherkassky1998learning} despite it being the most informative design.
% 2-LVL FACTORIAL
The 2-level factorial design provides an alternative with $2^d$ experiments (where $d$ is the number of independent variables). 
Drawbacks of 2-level factorial \gls{doe}s include no data about the inside of the search space and infeasibility for high dimensional problems.
In a full factorial or 2-level factorial design most experiments are redundant and most resources will be spent exploring high-order interaction effects\cite{gunst2009fractional}, which are often minimal to non-existent.
% RANDOM 
In order to overcome these obstacles a certain number of experiments can be chosen randomly from the search space. 
When a subset is chosen from the factorial design, it is called a fractional factorial design. 
% PLACKETT_BURMAN
The \gls{pb}\cite{vanaja2007design,miller2001using,wang1995hidden} design is a special case of 2-level fractional factorial design, 
where the number of needed experiments $n$ is $n<d+4$ 
(more precisely $n=(\lfloor d\div4\rfloor+1)\cdot4$, where $\lfloor x\rfloor$ denotes the floor function on $x$).
The \gls{pb} design ensures that each combination of levels for any pair of factors appears the same number of times. 
% HAMMERSLEY 
A drawback of 2-level factorial (incl. \gls{pb}) and random fractional designs is that the sample set is likely not evenly distributed across the search space\cite{viana2016tutorial}. 
The Hammersley design\cite{viana2016tutorial,diwekar1997efficient} is based on the Hammersley sequence and produces space filling data points. 
% LATIN HYPER CUBE
The Latin hypercube \gls{doe}\cite{viana2016tutorial,diwekar1997efficient} is a type of orthogonal \gls{doe}, 
which has the feature that each level for each variable will be tested only once. 
A Latin hypercube \gls{doe} can be also created such that data points distribute more uniformly over the search space. 
Latin hypercube \gls{doe}s are mainly used in computer simulations, which are purely deterministic and therefore are very precise.
%
%Continuous state spaces must be accommodated by arbitrary discretization.
%\cite{cohn1996neural}
\iffalse
%
\td{For any 2-level factorial DOE applies: "The effect of any factor main effect (A, B, C) or interaction (AB, AC, BC, ABC) is the difference of two averages, the average of the responses correpsonding to +1 levels and the average of the responses corresponding to -1 levels. " - from Gunst2009\cite{gunst2009fractional}}

@@

%%% FEATURE SELECTION 
% https://stats.stackexchange.com/questions/39243/how-does-one-interpret-svm-feature-weights
% gene (feature) selection with SVM https://link.springer.com/content/pdf/10.1023/A:1012487302797.pdf
% Intro to feature selection https://axon.cs.byu.edu/Dan/778/papers/Feature%20Selection/guyon2.pdf
\iffalse
%%% VALIDATION 
\subsection{Validation}
Validation is an important part of data analysis as it allows to estimate model errors. 
Often used methods are analysis of residual error, cross-validation and leave-one-out validation\cite{kartam1997artificial}.
\td{to do read kartam}
statistical methods to bewerten a model are: 
cross-validation, leave-one-out validation (special case of cross-validation), correlation coefficient for linear fits, mean square error, mean absolute erro
\td{cite from Hawkins 2004\cite{hawkins2004overfitting}}

### 2024-08-11 doc/Thesis/chapter/023-linreg-krr-svm.tex
start doing todo.md
nearly finish with todos from todo.md

%
\textbf{Linear regression} is one of the simplest methods to predict data. 
It persuades by its computational simplicity and easy interpretation. 
%Let's take this equation
\begin{equation}
\mathbf{y} = \mathbf{X} \mathbf{k} +c 
\end{equation}
Where $\mathbf{y} \in \mathbb{R}^n$ ($n$ is the number of data points) is a vector of dependent variables which shall be predicted and $\mathbf{X}\in \mathbb{R}^{n\times d}$ ($d$ is the number of independent variables) is a matrix of independent variables. 
The parameters $\mathbf{k} \in \mathbb{R}^d$ and $c \in \mathbb{R}$ are chosen by minimizing an objective function (loss function in \gls{ml} jargon).
A typical function to minimize are the $L_1$ and $L_2$ error, i.e. \gls{mae} and \gls{mse}, respectively.
\begin{align}
L_1&= \sum_i |y_i - \hat y_i| \\%= \sum_i \sum_j |y_i - x_{i,j}\cdot k_i+c| \\
L_2&= \sum_i(y_i - \hat y_i)^2 %= \sum_i \sum_j (y_i - x_{i,j}\cdot k_i+c)^2
\end{align}
%

\textbf{Ridge Regression} is like linear regression with an extra term which penalizes steep regression functions.
The extra term reduces overfitting and is scaled by a correction parameter~$\alpha$. 
The larger $\alpha$ is, the larger is the regularization and the flatter is the regression function. 
When $\alpha$ tends to infinity, we get an intercept-only model.
When $\alpha$ is zero, the regularized loss function becomes the $L_2$ loss function.
%\td{A critique of ridge regression is that all the variables tend to end up in the model. The model only shrinks the coefficients.}
\begin{equation}
    L_{RR} = L_2 + \alpha \sum |k_i| = \sum(y_i - \hat y_i)^2 + \alpha \sum |k_i| 
\end{equation}

%\textbf{Kernel Ridge Regression}
%%%%%%%%%%%%%%%%%%%%%%%%%%% { Kernel Ridge Regression } KRR %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\textbf{\Gls{krr}} combines ridge regression with the kernel method. 
A~kernel transforms data in such a way that a linear hyperplane (a point in one dimension, a line in two dimensions, 
a plane in three dimensions) can fit data in regression problems or separate the data in classification problems without actually doing the transformation for every data point, which lowers computational costs.
A kernel is some kind of similarity measure 
which fulfills the requirements of non-negativity, symmetry and linearity\cite{rupp2015machine}.
%\enlargethispage{\baselineskip}
%\looseness=-1
The following equations show definitions for 
linear~(eq.~\ref{eq:lin-kernel}), 
polynomial (eq.~\ref{eq:pol-kernel}), 
sigmoidal (eq.~\ref{eq:sig-kernel}) and 
radial basis functional kernel~(eq.~\ref{eq:rbf-kernel}), with $\gamma$ as fixed hyperparameter and $c_0$ as parameter to optimize.
A requirement for using kernels is having a dot product in the loss function. 
This can be accomplished by expressing $\mathbf{k}$ in terms of $\mathbf{X}$: $\mathbf{k}=\mathbf{X}^\top \mathbf{r}$
\cite{rudin2020least}.

>>>>

%
\textbf{Linear regression} is one of the simplest methods to predict data. 
It persuades by its computational simplicity and easy interpretation. 
%Let's take this equation
\begin{equation}
\mathbf{y} = \mathbf{X} \mathbf{k} +c 
\end{equation}
Where $\mathbf{y} \in \mathbb{R}^n$ ($n$ is the number of data points) is a vector of dependent variables which shall be predicted and $\mathbf{X}\in \mathbb{R}^{n\times d}$ ($d$ is the number of independent variables) is a matrix of independent variables. 
The parameters $\mathbf{k} \in \mathbb{R}^d$ and $c \in \mathbb{R}$ are chosen by minimizing an objective function (loss function in \gls{ml} jargon).
A typical function to minimize are the $L_1$ and $L_2$ error, i.e.\ \gls{mae} and \gls{mse}, respectively.
\begin{align}
L_1&= \sum_i |y_i - \hat y_i| \\%= \sum_i \sum_j |y_i - x_{i,j}\cdot k_i+c| \\
L_2&= \sum_i(y_i - \hat y_i)^2 %= \sum_i \sum_j (y_i - x_{i,j}\cdot k_i+c)^2
\end{align}
%

\textbf{Ridge regression} is like linear regression with an extra term, which penalizes steep regression functions.
The extra term reduces overfitting and is scaled by a correction parameter~$\alpha$. 
The larger $\alpha$ is, the larger is the regularization and the flatter is the regression function. 
When $\alpha$ tends to infinity, we get an intercept-only model.
When $\alpha$ is zero, the regularized loss function becomes the $L_2$ loss function.
%\td{A critique of ridge regression is that all the variables tend to end up in the model. The model only shrinks the coefficients.}
\begin{equation}
    L_{RR} = L_2 + \alpha \sum |k_i| = \sum(y_i - \hat y_i)^2 + \alpha \sum |k_i| 
\end{equation}

%\textbf{Kernel Ridge Regression}
%%%%%%%%%%%%%%%%%%%%%%%%%%% { Kernel Ridge Regression } KRR %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\textbf{\Gls{krr}} combines ridge regression with the kernel method. 
A~kernel transforms data in such a way that a linear hyperplane (a point in one dimension, a line in two dimensions, 
a plane in three dimensions) can fit data in regression problems or separate the data in classification problems without actually doing the transformation for every data point, which lowers computational costs.
A kernel is some kind of similarity measure, which fulfills the requirements of non-negativity, symmetry and linearity\cite{rupp2015machine}.
%\enlargethispage{\baselineskip}
%\looseness=-1
The following equations show definitions for 
linear~(eq.~\ref{eq:lin-kernel}), 
polynomial (eq.~\ref{eq:pol-kernel}), 
sigmoidal (eq.~\ref{eq:sig-kernel}) and 
radial basis functional kernel~(eq.~\ref{eq:rbf-kernel}), with $\gamma$ as fixed hyperparameter and $c_0$ as parameter to optimize.
A requirement for using kernels is having a dot product in the loss function. 
This can be accomplished by expressing $\mathbf{k}$ in terms of $\mathbf{X}$: $\mathbf{k}=\mathbf{X}^\top \mathbf{r}$
\cite{rudin2020least}.
